% Acknowledgements

\pdfbookmark[1]{Acknowledgements}{Acknowledgements} % Bookmark name visible in a PDF viewer

% \begin{flushright}{\slshape    
% Our most exciting discoveries come from studying anomalies. The once-in-1000 occurrence is worth getting detail on.} \\ \medskip
% --- {Michael J. Silverstein} 
% \end{flushright}
\bigskip

%----------------------------------------------------------------------------------------

\begingroup

\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter*{Acknowledgements}
I express my enormous gratitude to my advisor Odej Kao for including me in his research group and providing the opportunity to pursue and express my ideas. Odej provided the necessary freedom and trust so that I was able to identify problems and propose solutions in this thesis. Since the start of my doctoral studies, I had the pleasure to work with Jorge Cardoso from the Huawei Munich Research Center. His guidance throughout the years together with the huge experience and knowledge he shared with me were very beneficial. I thank Ingo Weber and Gjorgji Madjarov for accepting to review this thesis and providing valuable comments when I was allowed to present my results.

\bigskip

Special thanks to Jasmin Bogatinovski for the countless discussions on various topics, which not only improved my professional abilities but also empowered me to become a better version of myself. He critically read this thesis and helped me improve the presentation of my results with many practical suggestions and comments. 

\bigskip

I thank all my colleagues at the department of Distributed and Operating Systems for providing an enjoyable working atmosphere, sharing knowledge, and experience. Special thanks to Alexander Acker, Florian Schmidt, Ilya Verbitskiy, Lauritz Thamsen, and Tim Jungnickel. Enormous thanks to Jana Bechstein for her kindness and help with all administrative issues. I also had the luck to collaborate with Mihail Bogojeski and Lukas Ruff from the department of Machine Learning; thank you for the numerous insightful discussions. 

\bigskip

I thank all my friends and relatives for supporting me through all times. Particularly, I would like to thank Gjorgji I., Elena I., Petar T., Alisa K., Totikj, Elena T., Aleksandar P., Petar I., Nikola K., Marija Z., Irina I., Zoran S., Stefan G., Filip T., Stefan M., Davor T., and Petre P. for the constant help with their positive attitude.   

\bigskip

Ultimately, I thank my mother, Menka, for raising me with the right values and confidence to succeed on the path that I chose, my father Velche for his unconditional support, belief, and counsel, and my brother Zlatko for sharing his knowledge, which shaped my thinking and had the most beneficial influence in my life. I thank Nena for her love and patience, particularly during the most intense phases of my studies.

\endgroup